Doing Better Than UCT: \\ Rational Monte Carlo Sampling in Trees

### 2011-08-03 paper/rmcts/paper.tex
main results

\section{Introduction}

The simple regret of a sampling policy for the Multi-armed Bandit
Problem is the expected difference between the best expected reward
$\mu_*$ and the expected reward of the arm with the best sample mean
$\mu_j,\;\overline X_j=\max_i\overline X_i$:
\begin{equation}
\label{eq:simple-regret}
\IE[R]=\sum_{j=1}^K\Delta_j\Pr(\overline X_j=\max_i\overline X_i)
\end{equation}
where $\Delta_j=\mu_*-\mu_j$.

@@

@@

@@

\subsection{Simple regret in multi-armed bandits}

@@

\subsection{Monte Carlo tree search}

@@

\subsection{The sailing domain}

@@

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[scale=0.45]{rcq-size=6-nsamples=199.pdf}\\
    a. 199 playouts\\
    \vspace{1em}
    \includegraphics[scale=0.45]{rcq-size=6-nsamples=397.pdf}\\
    b. 397 playouts\\
  \end{minipage}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[scale=0.45]{rcq-size=6-nsamples=793.pdf}\\
    c. 793 playouts\\
    \vspace{1em}
    \includegraphics[scale=0.45]{rcq-size=6-nsamples=1585.pdf}\\
    d. 1585 playouts\\
  \end{minipage}
  \caption{The sailing domain, log vs. sqrt, $6\times 6$ lake}
  \label{fig:sailing-rcq-vs-factor}
\end{figure}
%
%\begin{figure}[ht]
%  \begin{minipage}[b]{0.5\linewidth} \centering
%    \includegraphics[scale=0.45,trim=0pt 0pt 0pt
%    0pt,clip]{sailing-size=13-factor=_5625.pdf}
%    {a. varying number of samples}
%  \end{minipage}
%  \begin{minipage}[b]{0.5\linewidth} \centering
%    \includegraphics[scale=0.45,trim=0pt 0pt 0pt
%    0pt,clip]{sailing-size=13-nsamples=240.pdf}
%    {b. varying exploration factor}
%  \end{minipage}
%  \caption{The sailing domain, a larger lake}
%  \label{fig:sailing-13}
%\end{figure}

\section{Summary and Future Work}

@@

\section{Derivations}

\subsection{Finite-time regret bounds}

@@

\subsubsection{$\varepsilon$-greedy}

@@

Number of times $\mu$ the $j$th arm is selected {\it randomly}:

\begin{equation}
\mu\triangleq\IE\left[T_j^R(n)\right]=\frac {n\varepsilon} K
\end{equation}

By choosing $x_0=\frac \mu 2$ and applying the Chernoff bound (\ref{eq:chernoff-bound}), obtain:

\begin{eqnarray}
\label{eq:pr-epsilon-greedy}
\Pr(\overline X_j=\max_i\overline X_i)&\le& 2\left(\frac {\mu}{2} e^{-\mu/8} + \frac {2e^{\Delta_j^2/2}}{\Delta_j^2}e^{-\Delta_j^2 \mu/4}\right)\nonumber\\
&\le&\left(\mu + \frac {4\sqrt e}{\Delta_j^2}\right)e^{-\Delta_j^2\mu/8}
\end{eqnarray}

@@

\begin{eqnarray}
  \IE r_\varepsilon&\le&\sum_{j=1}^K\Delta_j\left(\mu + \frac {4\sqrt e}
{\Delta_j^2}\right)e^{-\Delta_j^2\mu/8}\nonumber\\
&\le&K\left(\mu + \frac {4\sqrt
    e}{\Delta_{\min}}\right)e^{-\Delta_{\min}^2\mu/8}
\label{eq:deriv-epsilon-greedy-b}
\end{eqnarray}

\subsubsection{UCB($\alpha$)}

@@

\subsection{Asymptotic regret analysis}

@@

\subsubsection{$\varepsilon$-greedy}

A common technique \cite{Auer.ucb} to bound the probability of
selecting a non-optimal arm $j$ is to split the interval between
$\mu_j$ and $\mu_*$ at the middle and bound the probability of
selecting the $j$th arm by the sum of the probabilities that the samle
mean $\overline x_j$ of the $j$th arm is greater than $\mu_j+\frac
{\Delta_j} 2$ and the sample mean $\overline x_*$ of the optimal arm is
less than $\mu_*- \frac {\Delta_j} 2$. Each of the probabilities can
be bound using the Chernoff-Hoefding bound.

@@

\subsubsection{UCB($\alpha$)}

@@

Replacing the logarithm with a faster growing subpolynomial
function, like square root (UQB), results in a superpolynomial convergence
rate of the upper bound:
\begin{equation}
P_j^{uqb}=\mathbb{B}P_j^{uqb}\approx2\exp\left(-\beta\Delta_j^2\frac {\sqrt n}{\Delta_j^2}\right)
            =2\exp(-2\beta\sqrt n)
 \end{equation}
where $\beta$ is a constant depending on reward bounds.
Finite time convergence can still be shown similarly to UCB.

>>>>

\section{Introduction and definitions}

The Multi-armed Bandit is a set of $K$ arms. Each arm can be pulled
multiple times. When the $i$th arm is pulled, a random reward $X_i$
from an unknown stationary distribution is returned.  The reward is 
bounded between 0 and 1.

The simple regret of a sampling policy for the Multi-armed Bandit
Problem is the expected difference between the best expected reward
$\mu_*$ and the expected reward $\mu_j$ of the arm with the best sample mean
$\overline X_j=\max_i\overline X_i$:
\begin{equation}
\label{eq:simple-regret}
\IE[R]=\sum_{j=1}^K\Delta_j\Pr(\overline X_j=\max_i\overline X_i)
\end{equation}
where $\Delta_j=\mu_*-\mu_j$.
Strategies that minimize the simple reward are called pure exploration
strategies \cite{Bubeck.pure}.

Such strategies are used to select the best arm, and, by
extension, the best action in search, when every action in the
current world state is viewed as a different arm, and the best arm
must be chosen. Algorithm UCB that maximizes the accumulated reward
\cite{Auer.ucb} had been extended to tree search as UCT \cite{Kocsis.uct}.

UCT-driven search achieves good results in many search problems. At each
search step, the algorithm allocates Monte Carlo playouts to choose
the best action, thereby minimizing the simple regret. However, the
simple and the accumulated regret cannot be minimized simultaneously;
moreover, \cite{Bubeck.pure} shows that in many cases the smaller the
accumulated regret, the greater is the simple regret.

UCT performance can be improved by combining UCB with a sampling
scheme that minimizes the simple regret of selecting an action at the
current root node. Indeed, the algorithm must select an action with
the minimum regret \emph{given assumption that after performing the selected action
the algorithm performs optimally}, which corresponds to maximizing the
value of partial information \cite{Russell.aima}. Therefore, an
improved allocation scheme would
\begin{itemize}
\item maximize the value of partial
information by sampling actions to minimize the \textbf{simple} regret at the
current root node, and then 
\item minimize the \textbf{accumulated} regret of the playouts from the second
  step on to approach the optimum behavior. 
\end{itemize}

@@

The $\varepsilon$-greedy allocation scheme has exponential convergence
rate of the simple regret $\IE r$ (see Appendix~\ref{app:derivations-finite-eps}):
\begin{equation}
  \IE r_\varepsilon\le\sum_{j=1}^K\Delta_j\left(\frac {n\varepsilon} K + \frac {4\sqrt e}
{\Delta_j^2}\right)e^{-\Delta_j^2n\varepsilon/8K}
\end{equation}
while UCB exhibits polynomial convergence
rate (see Appendix~\ref{app:derivations-finite-ucb}):
\begin{equation}
\IE r_{ucb} \le 2\sum_{j=1}^K \Delta_jn^{\frac {-\alpha \Delta_j^2} 8}
\end{equation}
The best $\varepsilon$ is $\frac 1 2$, and for a large number of arms
the upper regret bound for the 1/2-greedy scheme approaches the square of the
upper regret bound for uniform sampling (see
Appendix~\ref{app:derivations-asym-eps}).

Another improved allocation scheme (UQB) for minimizing the simple
regret is obtained by substituting $\sqrt{\cdot}$ instead of $\log(\cdot)$ into
the formula for UCB and has a superpolynomial convergence rate
(see Appendix~\ref{app:derivations-asym-ucb}).

@@

By extension, a Monte Carlo tree search sampling scheme that selects
an action at the current root node according to an algorithm that
minimizes the simple regret, such as $1/2-greedy$ or $UQB$, and then
selects actions according to $UCB$, outperforms the UCT sampling
scheme that selects actions based on UCB only.

Such sampling scheme is significantly less sensitive to the tuning of
the empirical upper bound ($C_p$ in \cite{Kocsis.uct}) of UCT, since
the conflict \cite{Bubeck.pure} between the need for a larger value of $C_p$ on the first
step (simple regret) and a smaller value for the rest of the playout
(accumulated regret) is resolved. In fact, a sampling scheme that uses
UCB at all steps but a larger value of $C_p$ for the first step than
for the rest of the steps, outperforms UCT.

@@

The results were empirically verified on Multi-armed Bandit instances
(Section~\ref{seq:emp-mab}), on search trees
(Section~\ref{seq:emp-mcts}), and on the sailing domain
(Section~\ref{seq:emp-sailing}), as defined in \cite{Kocsis.uct}. The
experiments confirmed the hypotheses.

\subsection{Simple regret in multi-armed bandits}
\label{seq:emp-mab}

@@

\subsection{Monte Carlo tree search}
\label{seq:emp-mcts}

@@

\subsection{The sailing domain}
\label{seq:emp-sailing}

@@

\begin{figure}[ht]
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[scale=0.45]{rcq-size=6-nsamples=199.pdf}\\
    a. 199 playouts\\
    \vspace{1em}
    \includegraphics[scale=0.45]{rcq-size=6-nsamples=397.pdf}\\
    b. 397 playouts\\
  \end{minipage}
  \begin{minipage}[b]{0.5\linewidth}
    \centering
    \includegraphics[scale=0.45]{rcq-size=6-nsamples=793.pdf}\\
    c. 793 playouts\\
    \vspace{1em}
    \includegraphics[scale=0.45]{rcq-size=6-nsamples=1585.pdf}\\
    d. 1585 playouts\\
  \end{minipage}
  \caption{The sailing domain, log vs. sqrt, $6\times 6$ lake}
  \label{fig:sailing-rcq-vs-factor}
\end{figure}

%\section{Summary and Future Work}

@@

\section{Derivations}
\label{app:derivations}

\subsection{Finite-time regret bounds}
\label{app:derivations-finite}

@@

\subsubsection{$\varepsilon$-greedy}
\label{app:derivations-finite-eps}

@@

Number of times $\xi$ the $j$th arm is selected {\it randomly}:

\begin{equation}
\xi\triangleq\IE\left[T_j^R(n)\right]=\frac {n\varepsilon} K
\end{equation}

By choosing $x_0=\frac \xi 2$ and applying the Chernoff bound (\ref{eq:chernoff-bound}), obtain:

\begin{eqnarray}
\label{eq:pr-epsilon-greedy}
\Pr(\overline X_j=\max_i\overline X_i)&\le& 2\left(\frac {\xi}{2} e^{-\xi/8} + \frac {2e^{\Delta_j^2/2}}{\Delta_j^2}e^{-\Delta_j^2 \xi/4}\right)\nonumber\\
&\le&\left(\xi + \frac {4\sqrt e}{\Delta_j^2}\right)e^{-\Delta_j^2\xi/8}
\end{eqnarray}

@@

\begin{eqnarray}
  \IE r_\varepsilon&\le&\sum_{j=1}^K\Delta_j\left(\xi + \frac {4\sqrt e}
{\Delta_j^2}\right)e^{-\Delta_j^2\xi/8}\nonumber\\
&\le&K\left(\xi + \frac {4\sqrt
    e}{\Delta_{\min}}\right)e^{-\Delta_{\min}^2\xi/8}
\label{eq:deriv-epsilon-greedy-b}
\end{eqnarray}

\subsubsection{UCB($\alpha$)}
\label{app:derivations-finite-ucb}

@@

\subsection{Asymptotic regret analysis}
\label{app:derivations-asym}

@@

\subsubsection{$\varepsilon$-greedy}
\label{app:derivations-asym-eps}

A common technique \cite{Auer.ucb} to bound the probability of
selecting a non-optimal arm $j$ is to split the interval between
$\mu_j$ and $\mu_*$ at the middle and bound the probability of
selecting the $j$th arm by the sum of the probabilities that the samle
mean $\overline X_j$ of the $j$th arm is greater than $\mu_j+\frac
{\Delta_j} 2$ and the sample mean $\overline X_*$ of the optimal arm is
less than $\mu_*- \frac {\Delta_j} 2$. Each of the probabilities can
be bound using the Chernoff-Hoefding bound.

@@

\subsubsection{UCB($\alpha$)}
\label{app:derivations-asym-ucb}

@@

Replacing the logarithm with a faster growing subpolynomial
function, like square root (UQB), results in a superpolynomial convergence
rate of the upper bound:
\begin{equation}
P_j^{uqb}=\mathbb{B}P_j^{uqb}\approx2\exp\left(-\beta\Delta_j^2\frac {\sqrt n}{\Delta_j^2}\right)
            =2\exp(-2\beta\sqrt n)
 \end{equation}
where $\beta$ is a constant that depends on reward bounds only.
Finite time convergence can still be shown similarly to UCB.
